---
title: "Predict models to single scenario"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

-   [Introduction](#introduction)
-   [Predict selected models for a single scenario](#predict-selected-models-for-a-single-scenario)
    -   [Predict to SpatRaster](#predict-to-spatraster)
    -   [Predict to data.frame](#predict-to-data.frame)
    -   [Binarize models](#binarize-models)
    -   [Clamping variables](#clamping-variables)
    -   [No Extrapolation](#no-extrapolation)
    -   [Output Type](#output-type)
-   [Saving Predictions](#saving-predictions)

<hr>

## Introduction {#introduction}

Once selected models have been fit using `fit_selected()`, projections to single or multiple scenarios can be performed. The `predict_selected()` function is designed for projections to single scenarios.

To predict using the selected models, a `fitted_models` object is required. For detailed information on model fitting, please consult the vignette [Fit and Explore Selected Models](model_exploration.html). The `fitted_models` object generated in that vignette is included as an example dataset within the package. Let's load it.

```{r import maxnet calib, warning=FALSE}
#Load packages
library(kuenm2)
library(terra)

#Import calib_results_maxnet
data("fitted_model_maxnet", package = "kuenm2")
#Print calibration result
fitted_model_maxnet
```

To compare the results, let's import a `fitted_models` object generated using the GLM algorithm:

```{r import glm calib, warning=FALSE}
#Import calib_results_maxnet
data("fitted_model_glm", package = "kuenm2")
#Print calibration result
fitted_model_glm
```

<br>

# Predict Selected Models for a Single Scenario {#predict-selected-models-for-a-single-scenario}

To predict selected models for a single scenario, you need a `fitted_models` object and the corresponding predictor variables. These predictor variables can be provided as either a `SpatRaster` or a `data.frame`. The names of the variables (or columns in the `data.frame`) must precisely match those used for model calibration or those used when running PCA (if `do_pca = TRUE` was set in the `prepare_data()` function; see [Prepare Data for Model Calibration](prepare_data.html) for more details).

## Predict to SpatRaster {#predict-to-spatraster}

Let's use the same raster variables that were used to prepare the data and calibrate the models. These are included as example data within the package:

```{r Import raster layers}
# Import raster layers
var <- rast(system.file("extdata", "Current_variables.tif", package = "kuenm2"))
#Plot raster layers
plot(var)
```

Let's check which variables were used to calibrate our models. They are available in the `calibration_data` element of the object:

```{r check variables}
# Variables used to calibrate maxnet models
colnames(fitted_model_maxnet$calibration_data)

#Variables used to calibrate glm models
colnames(fitted_model_glm$calibration_data)
```

The first column, "pr_bg", indicates the presence (1) and absence (0) records, while the other columns represent the environmental variables. In this case, the variables are `bio_1`, `bio_7`, `bio_12`, `bio_15`, and `SoilType`. All these variables are present in the `SpatRaster` (`var`) we imported. Therefore, we can now predict our models to this raster. Let's begin by predicting the maxnet model:

```{r predict maxnet}
p_maxnet <- predict_selected(models = fitted_model_maxnet, 
                             raster_variables = var,
                             progress_bar = FALSE)
```

By default, the function computes consensus metrics (mean, median, range, and standard deviation) for each model across its replicates (if more than one model was selected), as well as a general consensus across all models. In this case, the output is a `list` containing `SpatRaster` predictions for each replicate, along with the consensus results for each model and the overall general consensus:

```{r check maxnet predictions}
#See objects in the output of predict_selected
names(p_maxnet)
```

Let's plot the general consensus:

```{r plot maxnet general}
plot(p_maxnet$General_consensus)
```

We can also plot the results for each replicate and the consensus for each model:

```{r plot models maxnet}
#Predictions for each replicate from model 159
plot(p_maxnet$Model_159$Replicates)

#Consensus across each replicate from model 159
plot(p_maxnet$Model_159$Model_consensus)
```

<br>

For comparison, let's predict the GLM model:

```{r}
# Predict glm model
p_glm <- predict_selected(models = fitted_model_glm, 
                          raster_variables = var,
                          progress_bar = FALSE)
#See selected models that were predicted
names(p_glm)

#Compare general consensus (mean) between maxnet and glm
par(mfrow= c(1, 2)) #Set grid to plot
plot(p_maxnet$General_consensus$mean, main = "Maxnet")
plot(p_glm$General_consensus$mean, main = "GLM")
on.exit() #Reinitiate grid
```

<br>

## Predict to data.frame {#predict-to-data.frame}

Instead of a `SpatRaster`, we can also predict the models to a `data.frame` that stores the variable values. To see an example, let's convert the raster variables `var` to a `data.frame`:

```{r var to df}
var_df <- as.data.frame(var)
head(var_df)
```

Note that each column stores the values for each variable. Let's predict our Maxnet models to this `data.frame`:

```{r predict to df}
p_df <- predict_selected(models = fitted_model_maxnet, 
                         raster_variables = var_df, #Now, a data.frame
                         progress_bar = FALSE) 
```

Now, instead of `SpatRaster` objects, the function returns `data.frame` objects with the predictions:

```{r predict df results}
#Results by replicate of the model 159
head(p_df$Model_159$Replicates)

#Consensus across replicates of the model 159
head(p_df$Model_159$Model_consensus)

#General consensus across all models
head(p_df$General_consensus)
```

<br>

## Binarize Models {#binarize-models}

The `fitted_models` object stores the thresholds that can be used to binarize the models into suitable and unsuitable areas. These thresholds correspond to the omission error rate used during model selection (e.g., 5% or 10%).

You can access the omission error rate used to calculate the thresholds directly from the object:

```{r omission error}
#Get omission error used to select models and calculate the thesholds
## For maxnet model
fitted_model_maxnet$omission_rate

## For glm model
fitted_model_glm$omission_rate
```

In both models, the omission error rate used to calculate the thresholds was 5%. This means that when the predictions are binarized, approximately 5% of the presence records used to calibrate the models will fall into cells classified as unsuitable.

The thresholds are summarized in two ways: the mean and median across replicates for each model, and the consensus mean and median across all selected models (when more than one model is selected). Let's check the thresholds for the general consensus:

```{r thresholds}
#For maxnet
fitted_model_maxnet$thresholds$consensus

#For glm
fitted_model_glm$thresholds$consensus
```

Let's use these thresholds to binarize the models (this functionality is only available when predicting to a `SpatRaster`):

```{r binarize models}
#Get the thersholds for models (general consensus)
thr_mean_maxnet <- fitted_model_maxnet$thresholds$consensus$mean #Maxnet
thr_mean_glm <- fitted_model_glm$thresholds$consensus$mean #glm

#Binarize models
mean_maxnet_bin <- (p_maxnet$General_consensus$mean > thr_mean_maxnet) * 1
mean_glm_bin <- (p_glm$General_consensus$mean > thr_mean_glm) * 1

#Compare results
par(mfrow= c(1, 2)) #Set grid to plot
plot(mean_maxnet_bin, main = "Maxnet")
plot(mean_glm_bin, main = "GLM")
on.exit() #Reinitiate grid
```

<br>

## Clamping Variables {#clamping-variables}

By default, predictions are performed with free extrapolation (`extrapolation_type = "E"`). This can be problematic when the peak of suitability occurs at the extremes of a predictor's range. For example, let's examine the response curve of the Maxnet model for `bio_7` (Temperature Annual Range):

```{r response bio_7}
response_curve(models = fitted_model_maxnet, variable = "bio_7", 
               extrapolation_factor = 1)
```

Note that higher suitability occurs at low values of the temperature range. However, the lower limit of the calibration data used to fit the models (dashed line) is at 15.7ºC. The premise that suitability will increase and stabilize at lower values of `bio_7` is an extrapolation of the model (the area to the left of the dashed line). It's possible that suitability could begin to decrease at extremely low values, rendering this extrapolation inaccurate, but the calibration data is insufficient for the model to predict this.

One way to address this is by clamping the variables. This means that all values outside the calibration range (both below the lower value and above the upper value) are set to the respective lower and upper limits of the calibration range. For example, in the calibration data for the Maxnet models, the lower and upper limits for `bio_7` are 15.7ºC and 23.3ºC, respectively:

```{r check lower and upper limits}
range(fitted_model_maxnet$calibration_data$bio_7)
```

To observe the effect of clamping this variable, let's create a hypothetical (and extreme) scenario where `bio_7` has extremely low values:

```{r create scenario}
#From bio_7, reduce values
new_bio7 <- var$bio_7 - 7
#Create new scenario
new_var <- var
#Replace bio_7 with new_bio7 in this scenario
new_var$bio_7 <- new_bio7

#Plot the differences
par(mfrow = c(1,2))
plot(var$bio_7, main = "Original bio_7")
plot(new_var$bio_7, main = "New bio_7")
on.exit() #Reinitiate grid
```

Let's predict the Maxnet models for this new scenario with both free extrapolation (`extrapolation_type = "E"`) and with clamped variables (`extrapolation_type = "EC"`):

```{r}
#Predict to hypothetical scenario with free extrapolation
p_free_extrapolation <- predict_selected(models = fitted_model_maxnet, 
                                         raster_variables = new_var, #New scenario
                                         consensus = "mean",
                                         extrapolation_type = "E", #Free extrapolation (Default)
                                         progress_bar = FALSE)

#Predict to hypothetical scenario with clamping
p_clamping <- predict_selected(models = fitted_model_maxnet, 
                               raster_variables = new_var, #New scenario
                               consensus = "mean",
                               extrapolation_type = "EC", #Extrapolation with clamping
                               progress_bar = FALSE)

#Get and see differences
p_difference <- p_free_extrapolation$General_consensus$mean - p_clamping$General_consensus$mean

#Plot the differences
par(mfrow = c(2,2))
plot(p_free_extrapolation$General_consensus$mean, main = "Free extrapolation",
     zlim = c(0, 1))
plot(p_clamping$General_consensus$mean, main = "Clamping",
     zlim = c(0, 1))
plot(p_difference, main = "Difference")
plot(new_bio7, main = "Hypothetical bio_7", type = "interval")
on.exit() #Reinitiate grid

```

Note that when we clamp the variables, regions with extremely low values of (the hypothetical) `bio_7` exhibit lower predicted suitabilities compared to when free extrapolation is allowed.

By default, when `extrapolation_type = "EC"` is set, all predictor variables are clamped. You can specify which variables to clamp using the `var_to_clamp` argument.

<br>

## No Extrapolation

A more rigorous approach is to **predict with no extrapolation**, where regions outside the limits of the calibration data are assigned a suitability value of 0. Let's predict the Maxnet models using the hypothetical scenario we created in the previous step to observe the difference:

```{r no extrapolation}
#Predict to hypothetical scenario with no extrapolation
p_no_extrapolation <- predict_selected(models = fitted_model_maxnet, 
                                       raster_variables = new_var, #New scenario
                                       consensus = "mean",
                                       extrapolation_type = "NE", #No extrapolation
                                       progress_bar = FALSE)
#Plot the differences
par(mfrow = c(2,2))
plot(p_free_extrapolation$General_consensus$mean, main = "Free extrapolation",
     zlim = c(0, 1))
plot(p_clamping$General_consensus$mean, main = "Clamping",
     zlim = c(0, 1))
plot(p_no_extrapolation$General_consensus$mean, main = "No extrapolation",
     zlim = c(0, 1))
plot(new_bio7, main = "Hypothetical bio_7", type = "interval")
on.exit() #Reinitiate grid
```

In this example, almost the entire predicted area shows zero suitability, except for a small patch. This occurred because, in this hypothetical scenario, nearly the entire region has `bio_7` values lower than those in the calibration data (which has a minimum of 15ºC). The only region where suitability is predicted to be greater than 0 is where the `bio_7` values fall within the limits of the calibration data.

By default, when `extrapolation_type = "NE"` is set, all predictor variables are considered for this process. You can specify a subset of variables to be considered for extrapolation using the `var_to_clamp` argument.

<br>

## Output Type

Maximum entropy models (maxnet) produce four different types of output for their predictions: raw, cumulative, logistic, and cloglog. These are described in [Merow et al. 2013](https://nsojournals.onlinelibrary.wiley.com/doi/10.1111/j.1600-0587.2013.07872.x) and [Phillips et al. 2017](https://nsojournals.onlinelibrary.wiley.com/doi/10.1111/ecog.03049).

All four output types are monotonically related. Therefore, rank-based metrics for model fit (e.g., omission rate and partial ROC) will be identical. However, the output types have different scaling, which leads to distinct interpretations and visually different prediction maps.

-   **Raw (or exponential) output** is interpreted as a Relative Occurrence Rate (ROR). The ROR sums to 1 when predicted to the calibration data.
-   **Cumulative output** assigns to a location the sum of all raw values less than or equal to the raw value for that location, and then rescales this to range between 0 and 100. Cumulative output can be interpreted in terms of an omission rate because thresholding at a value of *c* to predict a suitable/unsuitable cell will omit approximately *c*% of presences.
-   **Cloglog output (Default)** transforms the raw values into a scale of relative suitability ranging between 0 and 1, using a logistic transformation based on a user-specified parameter '$\tau$', which represents the probability of presence at 'average' presence locations. In this context, the tau value defaults to $\tau \approx 0.632$.
-   **Logistic output** is similar to Cloglog, but it assumes that $\tau = 0.5$.

Let's examine the differences between these four output types for Maxnet models:

```{r output types}
p_cloglog <- predict_selected(models = fitted_model_maxnet, raster_variables = var, 
                              type = "cloglog", progress_bar = FALSE)
p_logistic <- predict_selected(models = fitted_model_maxnet, raster_variables = var, 
                              type = "logistic", progress_bar = FALSE)
p_cumulative <- predict_selected(models = fitted_model_maxnet, raster_variables = var, 
                              type = "cumulative", progress_bar = FALSE)
p_raw <- predict_selected(models = fitted_model_maxnet, raster_variables = var, 
                              type = "raw", progress_bar = FALSE)

#Plot the differences
par(mfrow = c(2,2))
plot(p_cloglog$General_consensus$mean, main = "Cloglog (Default)",
     zlim = c(0, 1))
plot(p_logistic$General_consensus$mean, main = "Logistic",
     zlim = c(0, 1))
plot(p_cumulative$General_consensus$mean, main = "Cumulative",
     zlim = c(0, 1))
plot(p_raw$General_consensus$mean, main = "Raw",
     zlim = c(0, 1))
on.exit() #Reinitiate grid
```
<br>

# Saving Predictions

We can save the predictions to the disk by setting `write_files = TRUE`. When this option is enabled, you must provide a directory path in the `out_dir` argument.

If `raster_variables` is a `SpatRaster`, the function will save the output files as GeoTIFF (.tif) files. If `raster_variables` is a `data.frame`, the function will save the output files as Comma Separated Value (.csv) files.

```{r save, eval=FALSE}
p_save <- predict_selected(models = fitted_model_maxnet, 
                           raster_variables = var, 
                           write_files = TRUE, #To save to the disk
                           write_replicates = TRUE, #To save predictions for each replicate
                           out_dir = tempdir(), #A path to save the resuls (here, the temporary directory)
                           progress_bar = FALSE)
```

