---
title: "Project models to multiple scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

-   [Introduction](#introduction)
-   [Pre-processing raster predictors](#pre-processing-raster-predictors)
  -   [Organize and structure future climate variables from WorldClim](#organize-and-structure-future-climate-variables-from-worldClim)
    -   [Format for renaming](#format-for-renaming)
    -   [Fixed variables](#fixed-variables)
    -   [Organize and structure WorldClim files](#organize-and-structure-worldClim-files)
- [Preparation of data for model projections](#preparation-of-data-for-model-projections)
- [Project selected models to multiple scenarios](#project-selected-models-to-multiple-scenarios)
- [Import rasters resulting from projections](#import-rasters-resulting-from-projections)

<hr>

## Introduction {#introduction}

Once selected models have been fit using `fit_selected()`, projections to single or multiple scenarios can be performed. The `project_selected()` function is designed for projections to multiple scenarios.

To project using the selected models, a `fitted_models` object is required. For detailed information on model fitting, please consult the vignette [Fit and Explore Selected Models](/articles/model_exploration.html). The `fitted_models` object generated in that vignette is included as an example dataset within the package. Let's load it.

```{r import maxnet calib, warning=FALSE}
#Load packages
library(kuenm2)
library(terra)

#Import calib_results_maxnet
data("fitted_model_maxnet", package = "kuenm2")
#Print calibration result
fitted_model_maxnet
```

# Pre-processing raster predictors

While predicting models for a single scenario requires a single `SpatRaster` object containing the predictor variables (as detailed in [Predict models to single scenario](/articles/model_predictions.html)), projecting models to multiple scenarios necessitates a folder that stores the predictor variables for each scenario.

These folders must be organized in a specific hierarchical manner: a root directory should contain nested folders representing different scenarios, with the raster variables stored within. At the first level inside the root folder, subfolders should correspond to distinct time periods (e.g., future years like "2070" or "2100," or past periods such as "Mid-holocene" or "LGM"). Within each period folder, if applicable, you should include subfolders for each emission scenario (e.g., "ssp126", "ssp585"). Finally, within each emission scenario or time period folder, include a separate folder for each General Circulation Model (GCM) (e.g., "BCC-CSM2-MR", "MIROC6"). This structured organization enables the function to automatically access and process the data according to period, emission scenario, and GCM.

## Organize and structure future climate variables from WorldClim

The package provides a function to import future climate variables downloaded from WorldClim (version 2.1). This function renames the files and organizes them into folders categorized by period/year, emission scenario (Shared Socioeconomic Pathways; SSPs), and General Circulation Model (GCM). This simplifies the preparation of climate data, ensuring all required variables are properly structured for modeling projections.

To use this function, download the [future raster variables from WorldClim 2.1](https://worldclim.org/data/cmip6/cmip6climate.html) and save them all within the same folder. **DO NOT rename the files or the variables**, as the function relies on the patterns provided in the original files to work properly.

The package also provides an example of raw variables downloaded from WorldClim 2.1. This example includes bioclimatic predictions for the periods "2041-2060" and "2081-2100", for two SSPs (125 and 585) and two GCMs (ACCESS-CM2 and MIROC6), at 10 arc-minutes resolution.

```{r}
# See raster files with future predictors provided as example
# The data is located in the "inst/extdata" folder.
in_dir <- system.file("extdata", package = "kuenm2")
list.files(in_dir)
```

Note that all variables are in the same folder and retain the original names provided by WorldClim. You can download these variables directly from [WorldClim](https://worldclim.org/data/cmip6/cmip6climate.html) or by using the `geodata` R package:

```{r geodata,  eval=FALSE}
#Install geodata if necessary
if(!require("geodata")){
  install.packages("geodata")
}
#Load geodata
library(geodata)
#Create folder to save the raster files
#Here, in a temporary directory
geodata_dir <- file.path(tempdir(), "Future_worldclim")
dir.create(geodata_dir)
#Define GCMs, SSPs and time periods
gcms <- c("ACCESS-CM2", "MIROC6")
ssps <- c("126", "585")
periods <- c("2041-2060", "2061-2080")
#Create a grid of combination of periods, ssps and gcms
g <- expand.grid("period" = periods, "ssps" = ssps, "gcms" = gcms)
g #Each line is a specific scenario for future
#Loop to download variables for each scenario
lapply(1:nrow(g), function(i){
  cmip6_world(model = g$gcms[i], 
              ssp = g$ssps[i], 
              time = g$period[i], 
              var = "bioc", 
              res = 10, path = geodata_dir)
}) #It will take a while...
```

Let's check the variables inside the "geodata_dir" folder:

```{r check geodata folder, eval=FALSE}
list.files(geodata_dir, recursive = TRUE)
#> [1] "climate/wc2.1_10m/wc2.1_10m_bioc_ACCESS-CM2_ssp126_2041-2060.tif"
#> [2] "climate/wc2.1_10m/wc2.1_10m_bioc_ACCESS-CM2_ssp126_2061-2080.tif"
#> [3] "climate/wc2.1_10m/wc2.1_10m_bioc_ACCESS-CM2_ssp585_2041-2060.tif"
#> [4] "climate/wc2.1_10m/wc2.1_10m_bioc_ACCESS-CM2_ssp585_2061-2080.tif"
#> [5] "climate/wc2.1_10m/wc2.1_10m_bioc_MIROC6_ssp126_2041-2060.tif"    
#> [6] "climate/wc2.1_10m/wc2.1_10m_bioc_MIROC6_ssp126_2061-2080.tif"    
#> [7] "climate/wc2.1_10m/wc2.1_10m_bioc_MIROC6_ssp585_2041-2060.tif"    
#> [8] "climate/wc2.1_10m/wc2.1_10m_bioc_MIROC6_ssp585_2061-2080.tif" 
#> 
#> #Set climate as input directory
#> in_dir <- file.path(geodata_dir, "climate")
```
Now, we can organize and structure the files using the `organize_future_worldclim` function. 

### Format for renaming

An important argument is `name_format`, which defines the format for renaming variables. The names of the variables in the `SpatRaster` must precisely match those used for model calibration or when running PCA (if `do_pca = TRUE` was set in the `prepare_data()` function; see [Prepare Data for Model Calibration](/articles/prepared_data.html) for more details).

Therefore, if the variables used to calibrate the models are named "bio_1", "bio_2", etc., the variables in the future raster layers must also be named "bio_1", "bio_2", etc. However, if the variables have a different pattern, such as starting with uppercase letters and using zeros before single-digit numbers (e.g., "Bio_01", "Bio_02", etc.), they must be named "Bio_01", "Bio_02", etc. The function provides four options:

(1) `"bio_"`: Variables will be renamed to `bio_1`, `bio_2`, `bio_3`, `bio_10`, etc.
(2) `"bio_0"`: Variables will be renamed to `bio_01`, `bio_02`, `bio_03`, `bio_10`, etc.
(3) `"Bio_"`: Variables will be renamed to `Bio_1`, `Bio_2`, `Bio_3`, `Bio_10`, etc.
(4) `"Bio_0"`: Variables will be renamed to `Bio_01`, `Bio_02`, `Bio_03`, `Bio_10`, etc.

Let's check how the variables were named in our `fitted_model`:

```{r names in fitted}
fitted_model_maxnet$continuous_variables
```
The variables follows the standards of the first option (`"bio_"`). 

### Fixed variables

When predicting for other times, you can assume that some variables will be static (i.e., they remain unchanged in the projected scenarios). The `fixed_variables` argument allows you to append static variables alongside the bioclimatic variables.

Here, let's assume `soilType` will remain static in the future scenarios:

```{r static soil}
# Import raster layers (same used to calibrate and fit final models)
var <- rast(system.file("extdata", "Current_variables.tif", package = "kuenm2"))
#Get soilType
soiltype <- var$SoilType
```

### Organize and structure WorldClim files

Now, let's organize the WorldClim files with the `organize_future_worldclim()` function:

```{r organize_future_worldclim}
#Create folder to save structured files
out_dir_future <- file.path(tempdir(), "Future_raw") #Here, in a temporary directory
#Organize
organize_future_worldclim(input_dir = in_dir, #Path to the raw variables from WorldClim
                          output_dir = out_dir_future, 
                          name_format = "bio_", #Name format
                          fixed_variables = var$SoilType) #Static variables

# Check files organized
dir(out_dir_future, recursive = TRUE)
```

We can check the files structured hierarchically in nested folders using the `dir_tree()` function from the `fs` package:

```{r fs, eval = FALSE}
#Install package if necessary
if(!require("fs")){
  install.packages("fs")
}
dir_tree(out_dir_future)
#> Temp\RtmpkhmGWN/Future_raw
#> ├── 2041-2060
#> │   ├── ssp126
#> │   │   ├── ACCESS-CM2
#> │   │   │   └── Variables.tif
#> │   │   └── MIROC6
#> │   │       └── Variables.tif
#> │   └── ssp585
#> │       ├── ACCESS-CM2
#> │       │   └── Variables.tif
#> │       └── MIROC6
#> │           └── Variables.tif
#> └── 2081-2100
#>     ├── ssp126
#>     │   ├── ACCESS-CM2
#>     │   │   └── Variables.tif
#>     │   └── MIROC6
#>     │       └── Variables.tif
#>     └── ssp585
#>         ├── ACCESS-CM2
#>         │   └── Variables.tif
#>         └── MIROC6
#>             └── Variables.tif
```

After organizing variables, the next step is to create the `prepared_projection` object.

# Preparation of data for model projections

Now, let's prepare data for model projections across multiple scenarios, storing the paths to the rasters representing each scenario.

In contrast to `predict_selected()`, which requires a `SpatRaster` object, we need the paths to the folders where the raster files are stored. This includes the variables for the present time, which were used to calibrate and fit the models.

Currently, we only have the future climate files. The present-day predictor variables must reside in the same root directory as the processed future variables. Let's copy the rasters used for model calibration and fitting to this folder:

```{r copy present}
# Create a "Current_raw" folder in a temporary directory
# and copy the rawvariables there.
out_dir_current <- file.path(tempdir(), "Current_raw")
dir.create(out_dir_current, recursive = TRUE)

# Save current variables in temporary directory
terra::writeRaster(var, file.path(out_dir_current, "Variables.tif"))

#Check folder
list.files(out_dir_current)
```

Now, we can prepare the data for projections. In addition to storing the paths to the variables for each scenario, the function also verifies if all variables used to fit the final models are available across all scenarios. To perform this check, you need to provide either the `fitted_models` object you intend to use for projection or simply the variable names. We strongly suggest using the `fitted_models` object to minimize projection errors.

We also need to define the root directory containing the scenarios for projection (present, past, and/or future), along with additional information regarding time periods, SSPs, and GCMs.

```{r prepare projections}
# Prepare projections using fitted models to check variables
pr <- prepare_projection(models = fitted_model_maxnet,
                         present_dir = out_dir_current, #Directory with present-day variables
                         past_dir = NULL, #NULL because we won't project to the past
                         past_period = NULL, #NULL because we won't project to the past
                         past_gcm = NULL, #NULL because we won't project to the past
                         future_dir = out_dir_future, #Directory with future variables
                         future_period = c("2041-2060", "2081-2100"),
                         future_pscen = c("ssp126", "ssp585"),
                         future_gcm = c("ACCESS-CM2", "MIROC6"))
```

When we print the `projection_data` object, it summarizes all the scenarios we will predict and also shows the root directory where the predictor rasters are stored:

```{r print pr}
pr
```

If we check the structure of the `prepared_projection` object, we can see it's a list containing:

* Paths to all variables representing distinct scenarios in subfolders.
* The pattern used to identify the format of raster files within the folders (by default, `*.tif`).
* The names of the predictors.
* A list of class `prcomp` if a Principal Component Analysis (PCA) was performed on the set of variables with `prepare_data()`.

```{r str pr, eval = FALSE}
#Open prepared_projection in a new window
View(pr)
```

# Project selected models to multiple scenarios

After preparing the data, we can use the `project_selected()` function to predict the selected models across the multiple scenarios specified in `prepare_projections`:

```{r prepare_projections}
## Create a folder to save projection results
#Here, in a temporary directory
out_dir <- file.path(tempdir(), "Projection_results/maxnet")
dir.create(out_dir, recursive = TRUE)

## Project selected models to multiple scenarios
p <- project_selected(models = fitted_model_maxnet, 
                      projection_data = pr,
                      out_dir = out_dir, 
                      progress_bar = FALSE) #Do not print progress bar
```

The function returns a `model_projections` object. This object is similar to the `prepared_data` object, storing information on the predicted scenarios and the folder where the resulting projection rasters were saved.

```{r print model_projections}
print(p)
```

Note that the results were saved hierarchically in nested subfolders, each representing a distinct scenario. In the root directory, the function also saves a file named "Projection_paths.RDS", which is the `model_projections` object. This object can be imported into R using `readRDS(file.path(out_dir, "Projection_paths.RDS"))`.

```{r print results, eval=FALSE}
dir_tree(out_dir)
#> Temp\RtmpkhmGWN/Projection_results/maxnet
#> ├── Future
#> │   ├── 2041-2060
#> │   │   ├── ssp126
#> │   │   │   ├── ACCESS-CM2
#> │   │   │   │   ├── General_consensus.tif
#> │   │   │   │   ├── Model_159_consensus.tif
#> │   │   │   │   └── Model_189_consensus.tif
#> │   │   │   └── MIROC6
#> │   │   │       ├── General_consensus.tif
#> │   │   │       ├── Model_159_consensus.tif
#> │   │   │       └── Model_189_consensus.tif
#> │   │   └── ssp585
#> │   │       ├── ACCESS-CM2
#> │   │       │   ├── General_consensus.tif
#> │   │       │   ├── Model_159_consensus.tif
#> │   │       │   └── Model_189_consensus.tif
#> │   │       └── MIROC6
#> │   │           ├── General_consensus.tif
#> │   │           ├── Model_159_consensus.tif
#> │   │           └── Model_189_consensus.tif
#> │   └── 2081-2100
#> │       ├── ssp126
#> │       │   ├── ACCESS-CM2
#> │       │   │   ├── General_consensus.tif
#> │       │   │   ├── Model_159_consensus.tif
#> │       │   │   └── Model_189_consensus.tif
#> │       │   └── MIROC6
#> │       │       ├── General_consensus.tif
#> │       │       ├── Model_159_consensus.tif
#> │       │       └── Model_189_consensus.tif
#> │       └── ssp585
#> │           ├── ACCESS-CM2
#> │           │   ├── General_consensus.tif
#> │           │   ├── Model_159_consensus.tif
#> │           │   └── Model_189_consensus.tif
#> │           └── MIROC6
#> │               ├── General_consensus.tif
#> │               ├── Model_159_consensus.tif
#> │               └── Model_189_consensus.tif
#> ├── Present
#> │   └── Present
#> │       ├── General_consensus.tif
#> │       ├── Model_159_consensus.tif
#> │       └── Model_189_consensus.tif
#> └── Projection_paths.RDS
```

By default, for each scenario, the function computes consensus metrics (mean, median, range, and standard deviation) for each model across its replicates (if more than one model was selected), as well as a general consensus across all models.

The function accepts several other parameters that control the predictions in `predict_selected()`, such as the consensus to compute, the extrapolation type (free extrapolation (`E`), extrapolation with clamping (`EC`), and no extrapolation (`NE`)), variables to clamp, and the format of prediction values (`raw`, `cumulative`, `logistic`, or the default `cloglog`). For more details, consult the vignette for [Predict models to single scenario](/articles/model_predictions.html).

# Import rasters resulting from projections

The `model_projections` object stores only the paths to the resultant rasters. To import the results, we can use the `import_projections()` function. By default, it imports all consensus metrics ("median", "range", "mean", and "stdev") and all scenarios (time periods, SSPs, and GCMs) available in the `model_projections` object. Let's import the mean for all scenarios:

```{r import mean}
#Import mean of each projected scenario
p_mean <- import_projections(projection = p, consensus = "mean")
#Plot all scenarios
plot(p_mean, cex.main = 0.8)
```

Alternatively, we can import results from specific scenarios. For example, let's import the results only for the "2041-2060" time period under the SSP 126:

```{r specific scen}
p_2060_ssp126 <- import_projections(projection = p, consensus = "mean", 
                                    present = FALSE, #Do not import present projections
                                    future_period = "2041-2060",
                                    future_pscen = "ssp126")
#Plot all scenarios
plot(p_2060_ssp126, cex.main = 0.8)
```

With the `model_projections` object, we can compute changes in suitable areas between scenarios (see the `projection_changes` function), explore variance stemming from replicates, model parameterizations, and GCMs (see `projection_variability`), and perform analysis of extrapolation risks (see `projection_mop`). For more details, check the vignette on [Explore Variability and Uncertainty in Projections](/articles/variability_and_uncertainty.html).
