---
title: "7. Exploring Model Uncertainty and Variability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. Exploring Model Uncertainty and Variability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi=100,
  out.width = "95%"
)
```

# Summary

- [Description](#description)
- [Getting ready](#getting-ready)
- [Loading and preparing data](#loading-and-preparing-data)
- [Exploring variability](#exploring-variability)
    - [Importing results](#importing-results-1)
    - [Saving results](#saving-results-1)
- [Assessing extrapolation risks](#assessing-extrapolation-risks)
    - [MOP result options](#mop-result-options)
        - [Distances](#distances)
        - [Basic](#basic)
        - [Simple](#simple)
        - [Towards high and low ends](#towards-high-and-low-ends)
        - [Combined results](#combined-results)
    - [Handling values within range](#handling-values-within-range)
    - [MOP results and response curves](#mop-results-and-response-curves)
    - [Saving and importing results](#saving-and-importing-results)

<hr>

# Description

`kuenm2` has a set of functions that help explore variability and uncertainty of suitability results obtaining from projections to distinct scenarios. In short, the following analyses can be performed:

* **Explore variability** from replicates, model parameterizations, and General Circulation Models (GCMs) with `projection_variability()`.
* **Assess extrapolation risks** through analysis with `single_mop()` and `projection_mop()`.

<br>

# Getting ready

At this point it is assumed that `kuenm2` is installed (if not, see the [Main guide](../index.html)). Load `kuenm2` and any other required packages, and define a working directory (if needed). 

Note: functions from other packages (i.e., not from base R or `kuenm2`) used in this guide will be displayed as `package::function()`. 

```{r, results='hide', message=FALSE, warning=FALSE}
# Load packages
library(kuenm2)
library(terra)

# Current directory
getwd()

# Define new directory
#setwd("YOUR/DIRECTORY")  # uncomment and modify if setting a new directory

# Saving original plotting parameters
original_par <- par(no.readonly = TRUE)
```

<br>

# Loading and preparing data

When used in projects in which multiple projections were performed, these analyses require a `model_projections` object, which is the output of `project_selected()`. Let's load data and produce the required objects to continue with comparisons and variability and uncertainty estimations.

For more details about model projections, see ["Project Models to a Single Scenario"](model_predictions.html) and ["Project Models to Multiple Scenarios"](model_projections.html).

```{r create model_projections, warning=FALSE, results='hide'}
# Import calib_results_maxnet
data("fitted_model_maxnet", package = "kuenm2")

# Import path to raster files with future variables provided as example
# The data is located in the "inst/extdata" folder.
in_dir <- system.file("extdata", package = "kuenm2")

# Import raster layers (same used to calibrate and fit final models)
var <- rast(system.file("extdata", "Current_variables.tif", package = "kuenm2"))

# Get soilType
soiltype <- var$SoilType

# Organize and structure WorldClim files
# Create folder to save structured files
out_dir_future <- file.path(tempdir(), "Future_raw")  # Here, in a temporary directory

# Organize
organize_future_worldclim(input_dir = in_dir,  # Path to the raw variables from WorldClim
                          output_dir = out_dir_future, 
                          name_format = "bio_",  # Name format
                          static_variables = var$SoilType,  # Static variables
                          progress_bar = FALSE, overwrite = TRUE)

# Create a "Current_raw" folder in a temporary directory
# and copy the rawvariables there.
out_dir_current <- file.path(tempdir(), "Current_raw")
dir.create(out_dir_current, recursive = TRUE)

# Save current variables in temporary directory
terra::writeRaster(var, file.path(out_dir_current, "Variables.tif"), 
                   overwrite = TRUE)

# Prepare projections using fitted models to check variables
pr <- prepare_projection(models = fitted_model_maxnet,
                         present_dir = out_dir_current,  # Directory with present-day variables
                         future_dir = out_dir_future,  # Directory with future variables
                         future_period = c("2041-2060", "2081-2100"),
                         future_pscen = c("ssp126", "ssp585"),
                         future_gcm = c("ACCESS-CM2", "MIROC6"))

# Project selected models to multiple scenarios
## Create a folder to save projection results
# Here, in a temporary directory
out_dir <- file.path(tempdir(), "Projection_results/maxnet")
dir.create(out_dir, recursive = TRUE)

## Project selected models to multiple scenarios
p <- project_selected(models = fitted_model_maxnet, 
                      projection_data = pr,
                      out_dir = out_dir,
                      write_replicates = TRUE,
                      progress_bar = FALSE,  # Do not print progress bar
                      overwrite = TRUE)
```

<br>

# Exploring variability

When projecting models, predictions can vary according to different **replicates**, **model parameterizations**, and **GCMs**. The `projection_variability()` function quantifies variability from these sources, offering valuable insights into what makes predictions fluctuate the most and where.

The `projection_variability()` function requires a `model_projections` object, which is generated using `project_selected()`. By default, it uses the `median` consensus to summarize variance across selected models and GCMs. Alternatively, users can specify the `mean` to use it as a representative instead.

To be able to assess variability from replicates, make sure that `write_replicates = TRUE` is set when running `project_selected()`.

Below, we demonstrate how to calculate variance from these different sources (replicates, models, and GCMs) and save the results to the designated `out_dir` directory.

```{r compute variance}
# Create a directory to save results
v <- projection_variability(model_projections = p, write_files = TRUE,
                            output_dir = out_dir,
                            verbose = FALSE, overwrite = TRUE)
```

<br>

The output is an object named `variability_projections`, a list containing `SpatRaster` layers that represent the variance deriving from replicates, models, and GCMs for each scenario (including the present). These results can help detect areas where prediction variability is higher.

For example, for the present time scenario, the variance mainly comes from differences among replicates. 

```{r plot variability present, fig.width = 7, fig.height = 4}
# Variability for the present
terra::plot(v$Present, range = c(0, 0.15))
```

<br>

In the most pessimistic scenario (SSP5-8.5) for 2081–2100, variability is not too high and comes primarily from replicates and GCMs.

```{r plot variability future}
# Variability for Future_2081-2100_ssp585 
terra::plot(v$`Future_2081-2100_ssp585`, range = c(0, 0.1))
```

<br>

## Importing Results

Because `write_files = TRUE` was set, the `variability_projections` object includes the file path where the results were saved. You can use this path with the `import_results()` function to load the results when needed.

```{r root_directory, eval = FALSE}
# See the folder where the results were saved
v$root_directory
#> [1] "Temp/Projection_results/maxnet/variance"
```

<br>

As an example, we will import the results for the 2041–2060 period and the SSP1-2.6 scenario. In this scenario, the variability mainly originates from differences among the selected models (see below).

```{r import variability_projections}
# Importing results
v_2041_2060_ssp126 <- import_results(projection = v, 
                                     present = FALSE,  # Do not import results from the present time
                                     future_period = "2041-2060", 
                                     future_pscen = "ssp126")

# Plot
terra::plot(v_2041_2060_ssp126, range = c(0, 0.1))
```

<br>

## Saving results

The `variability_projections` object is a list that contains resulting `SpatRaster` layers (if `return_raster = TRUE`) and the directory path where the results were saved (if `write_files = TRUE`).

If the results were not saved to disk during the initial run, you can save the `SpatRaster` layers afterward using the `writeRaster()` function. For example, to save the variability map for one of the future scenarios:

```{r save raster variability, eval = FALSE}
writeRaster(v$`Future_2081-2100_ssp585`, 
            file.path(out_dir, "Future_2081-2100_ssp585.tif"))
```

<br>

If the results were saved to disk, the `variability_projections` object is automatically stored in a folder named *variance* within the specified `output_dir`. You can read this object to the R environment in future sessions using the `readRDS()` function:

```{r read variability projection}
v <- readRDS(file.path(out_dir, "variance/variability_projections.rds"))
```

<br>

This object can then be used to import the results with the `import_results()` function.

<br>

# Assessing extrapolation risks

When projecting models to new regions or time periods, it is common to encounter conditions that are non-analogous to those used to train models. 

For example, the maximum annual mean temperature (bio_1) in our model's calibration data was $22.7^\circ C$:

```{r see maximum temp}
max(fitted_model_maxnet$calibration_data$bio_1)
```

<br>

In contrast, in future scenarios, conditions are projected to become warmer. To illustrate this, let's import environmental variables from one of the GCMs (ACCESS-CM2) under the future scenario SSP5-8.5 and examine the maximum temperature:

```{r Highlight non-analogous}
# Import variables from the 2081-2100 period (SSP585, GCM MIROC6)
future_ACCESS_CM2 <- rast(file.path(out_dir_future,
                                "2081-2100/ssp585/ACCESS-CM2/Variables.tif"))

# Range of values in bio_1
terra::minmax(future_ACCESS_CM2$bio_1)

# Plot
terra::plot(future_ACCESS_CM2$bio_1, 
            breaks = c(-Inf, 22.7, Inf))  # Highlight regions with values above 22.7ºC
```

<br>

Note that temperatures are expected to exceed the current maximum temperature in most of the area under future conditions.

The functions `single_mop()` and `projection_mop()` compute the Mobility-Oriented Parity (MOP) metric ([Owens et al. 2013](https://doi.org/10.1016/j.ecolmodel.2013.04.011)) with recent updates implemented in the [mop package](https://CRAN.R-project.org/package=mop) ([Cobos et al. 2024](https://doi.org/10.21425/fob.17.132916)). This metric calculates dissimilarity between training and projection conditions, and detect areas with conditions outside training ranges. Areas identified outside training ranges and those with large dissimilarity (distances) values are where model predictions are a product of extrapolation. Depending on how the model is predicting outside ranges (see response curves), interpreting those results can be risky because of uncertainties in prediction behavior (hence the term *extrapolation risks*).

The MOP analysis requires the following inputs:

- Reference data: An object of class `fitted_models` returned by the `fit_selected()` function, or an object of class `prepared_data` returned by `prepare_data()`. These objects contain the environmental data used during model training.
- Data of interest for analysis: 
  - For `single_mop()`, a `SpatRaster` with the environmental variables representing a single scenario for model projections. 
  - For `projection_mop()`, an object of class `projection_data` returned by the `prepare_projection()` function, with the paths to raster layers representing environmental conditions of multiple scenarios for model projections.

Important note: Most likely, MOP needs to consider only the variables used in the selected models. For that reason, using a `fitted_models` object and setting `subset_variables = TRUE` is preferred.

By default, `projection_mop()` does not compute distances and performs a *basic* type of MOP, which highlights only regions with conditions outside training ranges. Alternatively, you can set:

- `type = "simple"` to compute an additional layer with the number of variables with values outside training conditions per location.
- `type = "detailed"` to add multiple layers that identify exactly which variables exhibit conditions outside training ranges, and how.
- `calculate_distance = TRUE` to compute multivariate distances from training to all conditions in the scenarios of projections.

Below, we perform a detailed MOP to identify areas with extrapolation risk in the future scenarios for which predictions were made:

```{r perform mop}
# Create a folder to save MOP results
out_dir_mop <- file.path(tempdir(), "MOPresults")
dir.create(out_dir_mop, recursive = TRUE)

# Run MOP
kmop <- projection_mop(data = fitted_model_maxnet, projection_data = pr, 
                       subset_variables = TRUE,
                       calculate_distance = TRUE,
                       out_dir = out_dir_mop, 
                       type = "detailed", 
                       overwrite = TRUE, progress_bar = FALSE)
```

<br>

This application returns a `mop_projections` object, which contains the paths to the directories where the results were saved. This object can be used with the `import_results()` function to load the results.

<br>

## MOP result options

The results of the MOP analysis provide multiple perspectives on extrapolation risks. Each component a different aspect of the dissimilarity between training and projection environmental conditions.

When importing results, you can specify the periods and scenarios (e.g., "2081-2100" and "ssp585"), as well as the type of MOP results to load. By default, all available MOP types are imported, these include: `distances` (if calculated), `basic`, `simple`, `towards_high_end`, `towards_low_end`, `towards_high_combined`, and `towards_low_combined`. 

Below, we explore all MOP results for the SSP5-8.5 scenario and the 2081–2100 period:

```{r import mop, warning=FALSE}
# Import results from MOP runs
mop_ssp585_2100 <- import_results(projection = kmop,
                                  future_period = "2081-2100", 
                                  future_pscen = "ssp585")

# See types of results
names(mop_ssp585_2100)
```

<br>

### Distances

The element `distances` represents Euclidean or Mahalanobis distances depending on what is defined in the argument `distance`. Large distance values indicate greater dissimilarity from training conditions, highlighting areas with high extrapolation risk.

```{r plot distance}
terra::plot(mop_ssp585_2100$distances)
```

<br>

### Basic

The element `basic` identifies areas where at least one environmental variable is outside training conditions. A value of '1' indicates the presence of non-analogous conditions in that specific area and scenario.

```{r plot basic}
terra::plot(mop_ssp585_2100$basic)
```

<br>

### Simple

The results in `simple` quantify the number of environmental variables in the projected area that outside training conditions.

```{r plot simple}
terra::plot(mop_ssp585_2100$simple)
```

<br>

### Towards high and low ends

The results in `towards_high_end` and `towards_low_end` identify which areas show non-analogous conditions for each of the variables independently. An important detailed considered here is that the results show conditions above the maximum (towards high) and below the minimum (towards low) values in training data.   

```{r plot towards end}
# Non-analogous conditions towards high values in the ACCESS-CM2 scenario
terra::plot(mop_ssp585_2100$towards_high_end$`Future_2081-2100_ssp585_ACCESS-CM2`)

# Non-analogous conditions towards low values in the MIROC6 scenario
terra::plot(mop_ssp585_2100$towards_low_end$`Future_2081-2100_ssp585_ACCESS-CM2`,
     main = names(mop_ssp585_2100$towards_low_end$`Future_2081-2100_ssp585_ACCESS-CM2`))
```

<br>

### Combined results

The `towards_high_combined` and `towards_low_combined` combine the previous results but keep the identity of the variables detected outside ranges. Specifically, `towards_high_combined` identifies variables with values exceeding those observed in training data, whereas `towards_low_combined` shows variables with values below the training range.

```{r plot towars combined}
# Variables with conditions above training range
terra::plot(mop_ssp585_2100$towards_high_combined)

# Variables with conditions below training range
terra::plot(mop_ssp585_2100$towards_low_combined)
```

<br>

## Handling values within range

By default, the functions that run the MOP assign `NA` to cells with values within the training range. Alternatively, you can assign a value of 0 to these cells by setting `na_in_range = FALSE` when running `projection_mop()`.

```{r mop zero}
# Create a folder to save MOP results, now assigning 0 to cells within the range
out_dir_mop_zero <- file.path(tempdir(), "MOPresults_zero")
dir.create(out_dir_mop_zero, recursive = TRUE)

# Run MOP
kmop_zero <- projection_mop(data = fitted_model_maxnet, projection_data = pr, 
                            subset_variables = TRUE, 
                            na_in_range = FALSE,  # Assign 0 to cells within range
                            calculate_distance = TRUE,
                            out_dir = out_dir_mop_zero, 
                            type = "detailed", 
                            overwrite = TRUE, progress_bar = FALSE)
```

<br>

Let's explore how this setting affects the *simple* and *detailed* MOP outputs:

```{r check difference zero mop, fig.width = 7, fig.height = 4}
# Import MOP for the scenario ssp585 in 2081-2100
mop_ssp585_2100_zero <- import_results(projection = kmop_zero,
                                       future_period = "2081-2100", 
                                       future_pscen = "ssp585")

# Compare with the MOP that assigns NA to cells within the calibration range
# Simple MOP
terra::plot(c(mop_ssp585_2100$simple$`Future_2081-2100_ssp585_ACCESS-CM2`, 
              mop_ssp585_2100_zero$simple$`Future_2081-2100_ssp585_ACCESS-CM2`),
            main = c("Within range as NA", "Within range as 0"))

# Detailed MOP
terra::plot(c(mop_ssp585_2100$towards_high_combined$`Future_2081-2100_ssp585_ACCESS-CM2`, 
              mop_ssp585_2100_zero$towards_high_combined$`Future_2081-2100_ssp585_ACCESS-CM2`),
            main = c("Within range as NA", "Within range as 0"),
            plg=list(cex=0.6))

```

<br>

## MOP results and response curves

While MOP helps us identify areas with non-analogous environmental conditions, whether risks of extrapolation exist depend on how model is behaving under non-analogous conditions. Model response curves are of great help to check model behavior under distinct conditions. 

For example, consider a future scenario represented by the ACCESS-CM2 GCM and the SSP5-8.5 for 2081–2100. Here, values of bio_1 (Annual Mean Temperature), bio_12 (Annual Precipitation), and bio_15 (Precipitation Seasonality) exceed the upper limits observed in training data (see below).

```{r plot towards high values}
# Non-analogous conditions towards high values in the ACCESS-CM2 scenario
terra::plot(mop_ssp585_2100$towards_high_combined$`Future_2081-2100_ssp585_ACCESS-CM2`)
```

<br>

Now, let's check model response curves for these variables. To better visualize how the model responds to environmental conditions in this future scenario, we can set the plotting limits using this scenario’s layers as `new_data`:

```{r response curves towards high, fig.height=3.8}
# Plotting response curves to check extrapolation towards the high end of variables
par(mfrow = c(1,3))  # Set plot grid
response_curve(models = fitted_model_maxnet, variable = "bio_1", 
               new_data = future_ACCESS_CM2)
response_curve(models = fitted_model_maxnet, variable = "bio_12", 
               new_data = future_ACCESS_CM2)
response_curve(models = fitted_model_maxnet, variable = "bio_15", 
               new_data = future_ACCESS_CM2)
```

<br>

In the response curves for bio_1, bio_12, and bio_15, higher values correspond to lower suitability, reaching zero near the upper limit of training data (indicated by a dashed line). Beyond this upper limit, suitability remains close to zero, thus, we are not concerned about extrapolation risks towards higher values for these variables.

Next, let's investigate the variables with values below the lower limit of training data:

```{r plot towards low values, fig.width = 7, fig.height = 4}
# Non-analogous conditions towards low values in the ACCESS-CM2 scenario
par(mfrow = c(1, 2))  # Set grid
terra::plot(mop_ssp585_2100$towards_low_combined$`Future_2081-2100_ssp585_ACCESS-CM2`)

## It's bio 7. Plot response curve:
response_curve(models = fitted_model_maxnet, variable = "bio_7", 
               new_data = future_ACCESS_CM2)
```

<br>

In some regions of the projection scenario, `bio_7` (Temperature Annual Range) exhibits values *below the lower limit of the calibration data*. The response curve indicates that suitability continues to increase when extrapolating to these lower values. This is a situation of concern due to *extrapolation risks* because we are *uncertain* about when suitability should start to decrease.

This example highlights why we strongly recommend interpreting MOP results alongside the response curves.

<br> 

```{r par_reset}
# Reset plotting parameters
par(original_par) 
```

<br>

## Saving and importing results

When `projection_mop()` is run, it automatically saves the `mop_projections` object as an RDS file in `out_dir`. You can load this object in R at any time using the `readRDS()` function:

```{r save mop rds}
# Check for RDS files in the directory where we saved the MOP results
list.files(path = out_dir_mop, pattern = "rds")

# Import the mop_projections file
kmop <- readRDS(file.path(out_dir_mop, "mop_projections.rds"))
```
